Linear and Differential Cryptanalysis of DES

### 2009-01-29 trunk/talk/draft1.tex

\section{Introduction}
\begin{frame}{Introduction}
\begin{block}{Quack!}
Introduction to DES and the two attacks
\end{block}
\end{frame}

\subsection{Data Encrytption Standard}
\begin{frame}
DES structure

\begin{frame}
DES pieces

\begin{frame}
DES mathematics

\subsection{Linear Cryptanalysis}
\begin{frame}
Vulnerabilities of DES

@@

\subsection{Differential Cryptanalysis}
\begin{frame}
Vulnerabilities of DES

@@

\section{Mathematical Framework}
\begin{frame}
Mathematical description of the attacks

@@

\subsection{Simplification WLOG}
\begin{frame}
Reducing DES down to SPN

We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis. We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.

\begin{itemize}
\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
\item Each round consists of a substitution and a permutation (much like in DES)
\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks. Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box. We use the same S-box throughout. 
\item The permutation is applied to all 16-bits of the round following the substitution.
\item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
\item The SPN cipher we will look at will consist of 4 rounds.
\end{itemize}

\begin{frame}
SPN Visually
%\pgfimage[scale=0.75]{SPN.PNG}
\end{frame}

\begin{frame}
Introducing SPN mathematically

\subsection{Linear Attack on SPN}
\begin{frame}
\large{Basic math concepts behind the Linear attack:}\\
\vspace{5 mm}
Linear cryptanalysis tries to take advantage of high probability occurrences of  linear expressions involving plaintext bits, ciphertext bits and subkey bits.  It is a known plaintext attack.\\
\vspace{5 mm}
Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
\end{frame}

\begin{frame}
Intro to linear / affine approximation ...

\begin{frame}
Linear characteristic / approximation of SBox ...

\begin{frame}
Piling Up lemma ...

@@

\subsection{Differential Attack on SPN}
\begin{frame}
Basic math concepts behind Differential attack ...

@@

\subsection{Actual Attacks}
\begin{frame}
Demo of attacks on SPN ... Runtime comparisons with Brute Force ...

@@

\begin{frame}
Break ... After coming back from the break, we should refresh the basic ideas behind the attacks once again ...

\section{Attacks on DES}
\begin{frame}
Back to DES ... Pull back from SPN ...

@@

\section{Conclusion}
\begin{frame}
Concluding remarks about the project ...

@@

\subsection{Take home Lesson}
\begin{frame}
If you are mounting the attacks ... To Do / Not To Do ...

>>>>

\section{Introduction}

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
A quick review of DES
\end{beamercolorbox}
\end{frame}

\subsection{Data Encrytption Standard}
\begin{frame}
\begin{columns}
\begin{column}{0.6\textwidth}
\begin{block}{DES Function}
\[ DES : (\mbox{Plaintext }(P), \mbox{ Key }(K))  \mapsto \mbox{Cipher }(C) \]
\end{block}

\begin{block}{DES Unit Blocks}
\begin{align*}
\sigma_{i+1} : \: & (0,1)^{64} \rightarrow (0,1)^{64} \\
& (L_i, R_i) \mapsto (L_i \oplus f(R_i,K_{i+1}), R_i) \\
\tau : \: & (0,1)^{64} \rightarrow (0,1)^{64} \\
& (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})
\end{align*}
\end{block}

\begin{itemize}
\item{Input for round $i+1$: $(L_i,R_i)$}
\item{Subkey for round $i+1$: $K_{i+1}$}
\item{Feistel function ($f$)}
\item{16-rounds for $i = 0, 1, ..., 15$}
\end{itemize}

\end{column}
\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{des_algo.jpg}
\caption{DES Algorithm}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\begin{columns}
\begin{column}{0.45\textwidth}
\begin{block}{Feistel Function}
\begin{align*}
f : \: & (0,1)^{32} \times (0,1)^{48} \rightarrow (0,1)^{32} \\
& (R,K)  \mapsto P (S_{Box}(E(R) \oplus K)) 
\end{align*}
\end{block}

\begin{itemize}
\item{Right half of the plaintext ($R$)}
\item{Expansion function \\ $E : (0,1)^{32} \rightarrow (0,1)^{48}$}
\item{Key for the round ($K$)}
\item{Confusion function (S-Boxes) \\ $S_i: (0,1)^6 \rightarrow (0,1)^4$}
\item{Diffusion function \\ $P : (0,1)^{32} \rightarrow (0,1)^{32}$}
\end{itemize}
\end{column}
\begin{column}{0.55\textwidth}
\begin{figure}
\includegraphics[totalheight=0.6\textheight]{des_feistel.jpg}
\caption{Feistel Round for DES}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
Mathematically, the different pieces of DES are:
\begin{itemize}
\item{Initial Permutation $IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
\item{DES Unit Block functions $\sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
\item{Transpositions $\tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
\item{Inverse Initial Permutation $IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
\end{itemize}

\begin{block}{Algebraic Representation of DES}
\begin{align*}
C &= DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P) \\
P &= DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)
\end{align*}
\end{block}

Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$
\end{frame}

\subsection{Linear Cryptanalysis}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Basic Idea of Linear Cryptanalysis
\end{beamercolorbox}
\end{frame}

@@

\subsection{Differential Cryptanalysis}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Basic Idea of Differential Cryptanalysis
\end{beamercolorbox}
\end{frame}

@@

\section{Mathematical Framework}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Dive into the Mathematics
\end{beamercolorbox}
\end{frame}

\subsection{Simplification WLOG}

\begin{frame}
\begin{columns}
\begin{column}{0.6\textwidth}
\begin{figure}
\includegraphics[totalheight=0.55\textheight]{des_feistel.jpg}
\end{figure}
Let us simplify DES by
\begin{itemize}
\item{Reducing the bitlength from $64$ to $16$}
\item{Removing expansion function $E(R)$}
\item{Taking identical S-Boxes of size $4$ bits}
\item{Repeating over less rounds}
\end{itemize}
\end{column}

\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.9\textheight]{spn.jpg}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Simple Substitution-Permutation Network
\end{beamercolorbox}
\end{frame}

\begin{frame}
We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis. We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.

\begin{itemize}
\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
\item Each round consists of a substitution and a permutation (much like in DES)
\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks. Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box. We use the same S-box throughout.
\item The permutation is applied to all 16-bits of the round following the substitution.
\item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
\item The SPN cipher we will look at will consist of 4 rounds.
\end{itemize}
\end{frame}

\begin{frame}
SPN S-Box and Permutation:\\
\vspace{5mm}
SPN uses a single 4-bit S-Box that has the following structure:
\begin{figure}
\includegraphics[width=0.9\textwidth]{spn_sbox.jpg}
\end{figure}

\vspace{5mm}
And the following 16-bit permutation:
\begin{figure}
\includegraphics[width=0.9\textwidth]{spn_perm.jpg}
\end{figure}

\vspace{5mm}

@@

\subsection{Linear Attack on SPN}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Mathematics of Linear Cryptanalysis
\end{beamercolorbox}
\end{frame}

\begin{frame}
\large{Basic Definitions for the Linear attack:}\\
\vspace{5 mm}
Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits. It is a known plaintext attack.\\
\vspace{5 mm}
Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
\end{frame}

\begin{frame}
Linear and Affine approximation of S-Box:\\
\vspace{5mm}
How do we come up with the desired expression for the entire cipher? We start by looking at the only non-linear component, the S-Box.\\
\vspace{5mm}
To find the linear/afine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
\[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box. \\
\vspace{5mm}
Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions. (this is for a 4-bit S-box)
\end{frame}

\begin{frame}
S-Box approximatin example:\\
\vspace{5mm}
We can take the following expression as an example:
\[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.\\
\vspace{5mm}
Similarly the following expression
\[ X_3 \oplus X_4 = Y_1 \oplus Y_4 \]
Has a bias of $-3/8$, which is higher in magnitude than the previous one. It is the magnitude of the bias that is important.
\end{frame}

\begin{frame}
The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.\\
\vspace{5mm}
\begin{figure}
\includegraphics[height=0.7\textheight]{spn_linear_approx.jpg}
\end{figure}
\end{frame}

\begin{frame}
What this means for 1 Round:\\
\vspace{5mm}
Note, from the previous table, that the expression $X_1 \oplus X_3 \oplus X_4 = Y_2$ has a bias of $+1/4$. \\
\vspace{5mm}
Note also that $U_1 = P \oplus K_1$. 
\vspace{5mm} We can now write down the following linear approximation across the 1st round of SPN:
\begin{eqnarray*}
V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \mbox{  $\leftarrow$ S-Box approximation above}\\
		  & = & (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8})
\end{eqnarray*}
This expression holds with probability of $3/4$ (bias of $+1/4$)
\end{frame}

\begin{frame}
We can get expressions that hold with some non-$1/2$ probability for every round. But we must somehow combine them to write an expression relating the plaintext and ciphertext bits $\Rightarrow$ We use the Piling Up Principle:\\
\vspace{5mm}
Note: \\
$X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
$X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$\\
\vspace{5mm}
Assume the following probability distribution
\[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
         p_1 & \mbox{$i = 0$};\\
         (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
\[ Pr(X_2 = i)  = \left\{ \begin{array}{ll}
         p_2 & \mbox{$i = 0$};\\
         (1 - p_2) & \mbox{$i = 1$}.\end{array} \right. \] 
\end{frame}

\begin{frame}
Piling Up Principle Continued:\\
\vspace{5mm}
Assuming that $X_1$ and $X_2$ are independant:
\[ Pr(X_1 = i, X_2 = j)  = \left\{ \begin{array}{ll}
         p_1 p_2 & \mbox{$i = 0, j = 0$};\\
         (1 - p_1) p_2 & \mbox{$i = 1, j = 0$} \\
			p_1 (1 - p_2) & \mbox{$i = 0, j = 1$} \\
			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence:
\begin{eqnarray*}
 Pr(X_1 \oplus X_2 = 0) & = & Pr(X_1=0,X_2=0) + Pr(X_1=1,X_2=1) \\
								& = & p_1p_2 + (1 - p_1)(1-p_2)
\end{eqnarray*}
If we write $p_k = 1/2 + \epsilon_k$ we have
\[ Pr(X_1 \oplus X_2 = 0) = 1/2 + 2\epsilon_1 \epsilon_2 \]
Hence the bias of $X_1 \oplus X_2 = 0$ is 
\[ 2 \epsilon_1 \epsilon_2 \]
That is, twice the product of the bias of the original expressions.
\end{frame}

\begin{frame}
Piling Up Principle Continued - The Piling Up Lemma:\\
\vspace{5mm}
We extend the idea on the previous slide to the general case: For $n$ independant random binary variables, $X_1, X_2, \dots, X_n$, 
\[ Pr(X_1 \oplus X_2 \cdots \oplus X_n = 0) = 1/2 + 2^{n-1} \prod_{i=1}^n \epsilon_i \]
In terms of bias,
\[ \epsilon_{1,2,\dots,n} = 2^{n-1} \prod_{i = 1}^n \epsilon_i \]
Where $\epsilon_{1,2,\dots,n}$ is the total bias of $X_1 \oplus X_2 \cdots \oplus X_n = 0$. \\
\vspace{5mm}
Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$.  
\end{frame}

@@

\subsection{Differential Attack on SPN}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Mathematics of Differential Cryptanalysis
\end{beamercolorbox}
\end{frame}

\begin{frame}
Basic math concepts behind Differential attack ...

@@

\subsection{Actual Attacks}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Let's get our hands dirty!
\end{beamercolorbox}
\end{frame}

\begin{frame}
Demo of attacks on SPN ... Runtime comparisons with Brute Force ...

@@

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
It's time for a Break!
\end{beamercolorbox}
\end{frame}

\section{Attacks on DES}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Do these work for DES?
\end{beamercolorbox}
\end{frame}

@@

\section{Conclusion}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Did we miss anything important?
\end{beamercolorbox}
\end{frame}

@@

\subsection{Take Home}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
`Take Home'
\end{beamercolorbox}
\end{frame}

\begin{frame}
If you are mounting the attacks ... To Do / Not To Do ...
